Staged typed type-preserving tagless ``interpreters'', \emph{finally}

### 2007-04-01 tagless/tagless-final.tex
Starting with Section 4

\authorinfo{Jacques Carette}
           {McMaster University}
           {carette@mcmaster.ca}
\authorinfo{Oleg Kiselyov}
           {FNMOC}
           {oleg@pobox.com}
\authorinfo{Chung-chieh Shan}
           {Rutgers University}
           {ccshan@cs.rutgers.edu}

@@

We focus on embedding a typed object language into a typed metalanguage.
Types are useful as usual because they document the programmer's
intentions and can be checked early by machine (see~\S\ref{encoding}).
To be concrete, we use the typed object language in
Figure~\ref{fig:object} throughout this paper.  We aim not just to
interpret the object language in the narrow sense but also for
compilation, partial evaluation, and other processing.

@@

In sum, by Church\hyp encoding terms using ordinary functions,
we achieve a tagless interpreter
for a typed object language in a metalanguage with a simple type
system \citep{hindley-principal,milner-theory}.  In this \emph{final}
rather than \emph{initial} approach, both kinds of run-time errors
in \S\ref{tagproblem} (applying a nonfunction and evaluating an open
term) are reported at compile time. Because the new interpreter uses no
universal type or
pattern matching, it never results in a run-time error and is in fact
total.  Because this totality is obvious not just to us but also to the
metalanguage implementation, we avoid the serious performance penalty
\citep{WalidICFP02} of error checking.
\ccshan{Need to cite Thiemann here and in related work?}

The interpreter above is wired directly into the
functions |b|, |lam|, |app|, and so on.  In the remainder of this paper, 
we explain how to abstract the interpreter so as
to ``evaluate'' the same term in many different
ways: interpretation, compilation, partial evaluation, size
computation, and so forth.

@@

Our contributions are
\begin{enumerate}
\item to evaluate, compile, and partially evaluate a typed object language
   in a typed metalanguage with a simple type system;
\item to check statically for well-typed and closed object programs;
\item clean, comparable implementations in MetaOCaml and Haskell;
\item a functor signature that encompasses all our interpreters, from
    evaluation and compilation to partial evaluation and CBN CPS
    transformation;
\item a clear way to extend the object language with more features;
\item and an approach to self-interpretation compatible with the
  above.  Self-interpretation turned out to be harder than expected.
\end{enumerate}
\jacques{can we state fairly clearly what type system we need?  I mean,
Haskell98(?) and MetaOCaml with (objects, polymorphic variants, ...).
With our let-polymorphism trick for self-interpretation, what do we really
need?}

@@

We can also compile our object language
into OCaml code using MetaOCaml's staging facilities. MetaOCaml
represents future-stage expressions of type |t| at the
present stage as values of type |('c,t) code| where |'c| is the
environment classifier \cite{WalidPOPL03,calcagno-ml-like}. Code values are created
by a \emph{bracket} form |.<e>.|, which specifies that the expression |e| is to be
evaluated at a future stage. The \emph{escape} |.~e| must occur
within a bracket and specifies that the expression |e| must be evaluated
at the current stage; its result, which must be a code value, is
spliced into the code being built by the enclosing bracket. The \emph{run} form |.! e| evaluates
the future-stage code value~|e| by compiling and linking it at run-time.
The bracket, escape, and run are akin to
quasi-quotation, unquotation, and |eval| of Lisp.

Inserting brackets and escapes appropriately into the evaluating
interpreter~|R| above yields the compiling interpreter~|C| below.
% this code does not have the 'sv parameter. It shows up later.
\begin{code}
module C = struct

@@

end
\end{code}
This is a completely straightforward staging of
|module R|.
This compiler simply produces
residual code with no optimization. For example, evaluating our |test1|
\begin{code}
let module E = EX(C) in E.test1 ()
\end{code}
gives the code value |.<(fun x_6 -> x_6) (true)>.|
of inferred type |('c, bool) C.repr|.

This compiler does not incur
any interpretive overhead: the
code produced for $\fun{x}x$ is simply |fun x_6 -> x_6| and contains no 
  call to the evaluator, unlike the recursive calls to |eval0| and
  |eval| in the |L e| lines in \S\ref{tagproblem}.
The resulting code obviously contains no tags and no pattern matching.

We have also implemented this compiler in Haskell. Since Haskell
has no (convenient, typed) staging facilities, we had to emulate
them. To be precise, we defined a data type |ByteCode| with
constructors such as |Var|, |Lam|, |App|, |Fix|, and |INT|.
Whereas our representation of object terms uses higher-order
abstract syntax, our bytecode uses integer-named
variables to be realistic. We then define 
\begin{code}
newtype C t = C (Int -> (ByteCode t, Int)) 
\end{code}
where |Int| is the counter for creating fresh variable
names. We define the compiler by making |C| an instance of the
class |Symantics|. The implementation\footnote{The implementation uses
GADTs because we also wanted to write a typed interpreter for 
the \textsf{ByteCode} \emph{datatype}.} is quite similar (but slightly more
verbose) than the corresponding MetaOCaml code given above. The
accompanying code gives the full details.

@@

Surprisingly, it is also possible to write a partial evaluator!  The
partial evaluator does have tags: \emph{binding time} or \emph{phase}
tags, i.e. whether a value is static or dynamic.
Pattern-matching on this phase tag is always exhaustive; we still
have no universal type and no tags for object types.  The partial
evaluator uses the previously mentioned compiler to ``interpret'' dynamic
terms, and uses the interpreter to reduce static terms.  In this sense, 
the partial evaluator is a modular extension to both the compiler and
the interpreter.

\subsection{The trouble with (opaque) functions}
\label{S:PE-problem}
We want to combine the interpreter |R| and the compiler |C| to produce
an (online) partial evaluator.  We defer the discussion of the 
choice of online/offline partial evaluation until after we have presented our
solution in~\ref{S:PE-solution}.
The partial evaluator (PE) is too an interpreter: it
evaluates a term either to an immediate (or, `static' or present-stage)
value or to a `dynamic', future-stage value, or `code' (of the sort
produced by the compiler C). The interpreter attempts to perform all
possible computations (i.e., computations whose operarands are
present-stage values) as the present-stage. Our first attempt, in
Haskell, is straightforward: we introduce
\begin{code}
data P1 t = S1 (R t) (C t) | E1 (C t)
\end{code}
as a `direct sum' of the interpreter |R| and the compiler |C|. The
datatype |P1| has two \emph{phase} (not type!) tags |S1| 
(for static) and |E1| for dynamic. The static value has both the
present-stage and future-stage alternatives. The dynamic value does
not have the present-stage component. We define
\begin{code}
abstr1 :: P1 t -> C t
abstr1 (S1 _ dyn) = dyn
abstr1 (E1 dyn)   = dyn
\end{code}
to extract the future-stage component. We then attempt to define the
interpreter |P1| compositionally -- and it works for the first-order
fragment of our DSL language.

\begin{code}
instance Symantics P1 where
    int  x  = S1 (int x) (int x)
    bool b  = S1 (bool b) (bool b)
    add (S1 n1 _) (S1 n2 _) = int (unR (add n1 n2))
    add e1 e2 = E1 (add (abstr1 e1) (abstr1 e2))
    -- mul and leq are analogous and elided
    if_ (S1 s _) et ef = if unR s then et else ef
    if_ eb et ef = 
       E1 (if_ (abstr1 eb) (abstr1 et) (abstr1 ef))
\end{code}
Integer and boolean literals are obviously immediate, present-stage
values. The result of adding two values is static only if both values
are. If so, we use the evaluator |R| to add the values; the result is
an essentially integer literal, which can be promoted to the future
stage. If at least one of the operands to |add| is dynamic, we promote
both operands to the future-stage and use the compiler |C| to
`compile' the code for adding the operand expressions.  Because the
static value includes as a component its future-stage variant, and
because we build both components together, we are able to get by
without the polymorphic `lift' operation (that is, lift the
computation of any type from the present-stage to the future
stage). We can do such a lifting only on type-by-type basis (for
booleans and for integers). The problem with the polymorphic lift is
that it inhibits true compilation, to assembler or other similar first
order target language with no polymorphic (run-time reified) types.
The representation of integers in the target language most likely
differs from the representation of booleans or functions,
etc. Therefore, without the intentional type analysis, the polymorphic
lift is not compilable.

@@

\subsection{The `final' solution}
\label{S:PE-solution}
Let us re-examone the problem in Section~\ref{S:PE-problem}. What we
would ideally like is to write
\begin{code}
data P t = S (repr_pe t) (C t) | E (C t)
\end{code}
where |repr_pe| is a type function, defined inductively
% inductively because P below depends on repr_pe
as 
\begin{code}
repr_pe Int    = Int
repr_pe Bool   = Bool
repr_pe (a->b) = P a -> P b
\end{code}
Although we can use typeclasses to define that type function
in Haskell, that is not portable to MetaOCaml. We notice however that
the type case alternatives above are all associated with the existing
methods of |Symantics| and so the case analysis is already present. 
A simple and more portable solution thus emerges: we bake |repr_pe| 
into the signature |Symantics|. Switching to MetaOCaml for
portability, we recall from Section~\ref{encoding} that |repr| type
had two arguments |type ('c,'dv) repr|. We add an extra argument,
|'sv|, which is the result of applying |repr_pe| function above to the
type |'dv|. The signature |Symantics| now reads
\begin{code}
module type Symantics = sig
  type ('c,'sv,'dv) repr
  val int  : int  -> ('c,int,int) repr
  val bool : bool -> ('c,bool,bool) repr
  val add  : ('c,int,int) repr -> 
      ('c,int,int) repr -> ('c,int,int) repr
  val mul  : ('c,int,int) repr -> 
      ('c,int,int) repr -> ('c,int,int) repr
  val leq  : ('c,int,int) repr -> 
      ('c,int,int) repr -> ('c,bool,bool) repr
  val if_  : ('c,bool,bool) repr ->
             (unit -> ('c,'sa,'da) repr) ->
             (unit -> ('c,'sa,'da) repr) -> 
             ('c,'sa,'da) repr 
  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr)
    -> ('c,(('c,'sa,'da) repr -> ('c,'sb,'db) repr),
           'da->'db) repr
  val app : ('c,(('c,'sa,'da) repr -> ('c,'sb,'db) repr),
             'da->'db) repr
    -> ('c,'sa,'da) repr -> ('c,'sb,'db) repr
  val fix : 
    (('c,(('c,'sa,'da) repr -> ('c,'sb,'db) repr) as 's,
         'da->'db) repr 
    -> ('c,'s,'da->'db) repr)  -> ('c,'s,'da->'db) repr
end;;
\end{code}
The interpreters |R|, |L| and |C| above only relied on the last
argument of the |repr| type, which is unchanged. And so is their
implementation; the newly added argument |'sv| will be phantom.
For example, we would now write the interpreter |C| as
\begin{code}
module C = struct
  type ('c,'sv,'dv) repr = ('c,'dv) code
  (* the rest is the same *)
\end{code}
When writing the partial evaluator however, we make use of the
type argument |'sv|. Here's the corresponding instance of the
Symantics:

>>>>

\begin{comment}
\authorinfo{Jacques Carette}
           {McMaster University}
           {carette@mcmaster.ca}
\authorinfo{Oleg Kiselyov}
           {FNMOC}
           {oleg@pobox.com}
\authorinfo{Chung-chieh Shan}
           {Rutgers University}
           {ccshan@cs.rutgers.edu}
\end{comment}
\authorinfo{}{}{}

@@

We focus on embedding a typed object language into a typed metalanguage.
Types are useful as usual because they document the programmer's
intentions and can be checked early by machine (see~\S\ref{encoding}).
To be concrete, we use the typed object language in
Figure~\ref{fig:object} throughout this paper.  We aim not just for
evaluation of object programs but also for
compilation, partial evaluation, and other processing.

@@

In sum, by Church\hyp encoding terms using ordinary functions,
we achieve a tagless evaluator
for a typed object language in a metalanguage with a simple type
system \citep{hindley-principal,milner-theory}.  In this \emph{final}
rather than \emph{initial} approach, both kinds of run-time errors
in \S\ref{tagproblem} (applying a nonfunction and evaluating an open
term) are reported at compile time. Because the new interpreter uses no
universal type or
pattern matching, it never results in a run-time error and is in fact
total.  Because this totality is obvious not just to us but also to the
metalanguage implementation, we avoid the serious performance penalty
\citep{WalidICFP02} of error checking.
\ccshan{Need to cite Thiemann here and in related work?}

The evaluator above is wired directly into the
functions |b|, |lam|, |app|, and so on.  In the rest of this paper, 
we explain how to abstract the interpreter so as
to process the same term in many other
ways: compilation, partial evaluation, size
computation, and so forth.

@@

Our contributions are
\begin{enumerate}
\item to evaluate, compile, and partially evaluate a typed object language
   in a typed metalanguage with a simple type system;
\item to check statically for well-typed and closed object programs;
\item clean, comparable implementations in MetaOCaml and Haskell;
\item a functor signature that encompasses all our interpreters, from
    evaluation and compilation to partial evaluation and CBN CPS
    transformation;
\item a clear way to extend the object language with more features;
\item and an approach to self\hyp interpretation compatible with the
  above.  Self\hyp interpretation turned out to be harder than expected.
\end{enumerate}
\jacques{can we state fairly clearly what type system we need?  I mean,
Haskell98(?) and MetaOCaml with (objects, polymorphic variants, ...).
With our let-polymorphism trick for self-interpretation, what do we really
need?}

@@

Besides immediate evaluation, we can compile our object language
into OCaml code using MetaOCaml's staging facilities. MetaOCaml
represents future-stage expressions of type |t| at the
present stage as values of type |('c,t) code| where |'c| is the
environment classifier \cite{WalidPOPL03,calcagno-ml-like}. Code values are created
by a \emph{bracket} form |.<e>.|, which specifies that the expression |e| is to be
evaluated at a future stage. The \emph{escape} |.~e| must occur
within a bracket and specifies that the expression |e| must be evaluated
at the current stage; its result, which must be a code value, is
spliced into the code being built by the enclosing bracket. The \emph{run} form |.!e| evaluates
the future-stage code value~|e| by compiling and linking it at run-time.
The bracket, escape, and run are akin to
quasi-quotation, unquotation, and |eval| of Lisp.

Inserting brackets and escapes appropriately into the
evaluator~|R| above yields the compiler~|C| below.
% this code does not have the 'sv parameter. It shows up later.
\begin{code}
module C = struct

@@

end
\end{code}
This is a completely straightforward staging of
|module R|.
This compiler simply produces
residual code with no optimization. For example, interpreting our |test1|
\begin{code}
let module E = EX(C) in E.test1 ()
\end{code}
gives the code value |.<(fun x_6 -> x_6) (true)>.|
of inferred type |('c, bool) C.repr|.

This compiler does not incur
any interpretive overhead: the
code produced for $\fun{x}x$ is simply |fun x_6 -> x_6| and does not
  call the interpreter, unlike the recursive calls to |eval0| and
  |eval| in the |L e| lines in \S\ref{tagproblem}.
The resulting code obviously contains no tags and no pattern matching.

We have also implemented this compiler in Haskell. Since Haskell
has no (convenient, typed) staging facilities, we had to emulate
them. To be precise, we defined a data type |ByteCode| with
constructors such as |Var|, |Lam|, |App|, |Fix|, and |INT|.
Whereas our representation of object terms uses higher-order
abstract syntax, our bytecode uses integer-named
variables to be realistic. We then define 
\begin{code}
newtype C t = C (Int -> (ByteCode t, Int)) 
\end{code}
where |Int| is the counter for creating fresh variable
names. We define the compiler by making |C| an instance of the
class |Symantics|. The implementation\footnote{The implementation uses
GADTs because we also wanted to write a typed interpreter for 
the \textsf{ByteCode} \emph{data type}.} is quite similar (but slightly more
verbose) than the corresponding MetaOCaml code given above. The
accompanying code gives the full details.

@@

Surprisingly, we can write a partial evaluator using the idea above,
namely to build object terms using ordinary functions rather than data
constructors.  We present this partial evaluator in a sequence of four
attempts, the last of which finally uses no universal type and no tags
for object types.  We then discuss residualization and binding-time
analysis.  Our partial evaluator is a modular extension to the evaluator
in~\S\ref{S:interpreter-RL} and the compiler in~\S\ref{compiler}, in
that it uses the former to reduce static terms and the latter to build
dynamic terms.

\ccshan{Move later:} The
partial evaluator does have tags: \emph{binding time} or \emph{phase}
tags, i.e. whether a value is static or dynamic.
Pattern-matching on this phase tag is always exhaustive.

\subsection{Avoiding polymorphic lift}

Roughly speaking, a partial evaluator interprets each object term either
to a static (present-stage) term, as produced by the evaluator~|R|, or
to a dynamic (future-stage) term, as produced by the compiler~|C|.  The
interpreter computes as much as possible statically.  Thus one might try
to define partial evaluation as the Haskell data type
\begin{code}
data P0 t = S0 (R t) | E0 (C t)
\end{code}
To extract a dynamic term from this type, we create the functions
\begin{code}
abstrInt :: P0 Int -> C Int
abstrInt (S0 r) = int (unR r)
abstrInt (E0 c) = c

abstrBool :: P0 Bool -> C Bool
abstrBool (S0 r) = bool (unR r)
abstrBool (E0 c) = c
\end{code}
We then start to define |P0| as an instance of the |Symantics| class.
\begin{code}
instance Symantics P0 where
  int  x = S0 (int x) (int x)
  bool x = S0 (bool x) (bool x)
  add (S0 e1) (S0 e2) = S0 (add e1 e2)
  add e1 e2 = E0 (add (abstrInt e1)
                      (abstrInt e2))
\end{code}
Integer and boolean literals are obviously immediate, present-stage
values. Addition yields a static term if and only if both operands
are static. If so, we use the evaluator~|R| to add the operands.
If not, we extract dynamic terms from the operands and add them
using the compiler~|C|.

Whereas |mul| and |leq| are as easy to define as |add|, we encounter
a problem with |if_|.  If the first argument to |if_| is a dynamic term
(type |C Bool|) and the second and third arguments are a static term
(type |R a|) and a dynamic term (type |C a|), then we need to convert
the static term to dynamic, but there is no polymorphic ``lift''
function |a -> C a| to send a value to the future stage.

Our |Symantics| class includes only separate lifting methods |bool| and
|int|, not a parametrically polymorphic lifting method, for good reason:
When compiling to a first-order target language such as machine code,
booleans, integers, and functions may well be represented differently.
Thus polymorphic lift cannot be compiled without intensional type
analysis.  To avoid the need for polymorphic lift, we turn to
a technique due to Asai \ccshan{cite?}: build a dynamic term alongside
every static term.

\subsection{Delaying binding-time analysis}
\label{S:PE-problem}

We switch to the Haskell data type
\begin{code}
data P1 t = P1 (Maybe (R t)) (C t)
\end{code}
so that a partially evaluated term |P1 t| always contains a dynamic
component and sometimes contains a static component.  The two
alternative constructors of a |Maybe| value, |Just| and |Nothing|,
tag each partially evaluated term with a phase: either present or
future.  This tag is not an object type tag: all pattern matching below
is exhaustive.  Because the future-stage component is always present, we
can now define the polymorphic function
\begin{code}
abstr1 :: P1 t -> C t
abstr1 (P1 _ dyn) = dyn
\end{code}
to extract it without requiring polymorphic lift into~|C|.  We then try
to define the interpreter |P1|---and get as far as the first-order
constructs of our object language, including |if_|.
\begin{code}
instance Symantics P1 where
  int  x = P1 (Just (int x)) (int x)
  bool b = P1 (Just (bool b)) (bool b)
  add (P1 (Just n1) _) (P1 (Just n2) _)
    = int (unR (add n1 n2))
  add e1 e2 = P1 Nothing
    (add (abstr1 e1) (abstr1 e2))
  -- mul and leq are analogous and elided
  if_ (P1 (Just s) _) et ef
    = if unR s then et else ef
  if_ eb et ef = P1 Nothing
    (if_ (abstr1 eb) (abstr1 et) (abstr1 ef))
\end{code}

@@

\subsection{The `final' solution}
\label{S:PE-solution}
Let us re-examone the problem in Section~\ref{S:PE-problem}. What we
would ideally like is to write
\begin{code}
data P t = S (repr_pe t) (C t) | E (C t)
\end{code}
where |repr_pe| is a type function, defined inductively
% inductively because P below depends on repr_pe
as 
\begin{code}
repr_pe Int    = Int
repr_pe Bool   = Bool
repr_pe (a->b) = P a -> P b
\end{code}
Although we can use typeclasses to define that type function
in Haskell, that is not portable to MetaOCaml. We notice however that
the type case alternatives above are all associated with the existing
methods of |Symantics| and so the case analysis is already present. 
A simple and more portable solution thus emerges: we bake |repr_pe| 
into the signature |Symantics|. Switching to MetaOCaml for
portability, we recall from Section~\ref{encoding} that |repr| type
had two arguments |type ('c,'dv) repr|. We add an extra argument,
|'sv|, which is the result of applying |repr_pe| function above to the
type |'dv|. Figure~\ref{fig:ocaml} shows the new signature |Symantics|.
\begin{figure*}
\begin{floatrule}
\begin{code}
module type Symantics = sig

  type ('c,'sv,'dv) repr

  val int : int  -> ('c,int,int) repr
  val bool: bool -> ('c,bool,bool) repr

  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x) -> ('c,'x,'da->'db) repr
  val app : ('c,'x,'da->'db) repr -> (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
  val fix : ('x -> 'x) -> (('c, ('c,'sa,'da) repr -> ('c,'sb,'db) repr, 'da->'db) repr as 'x)

  val add : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
  val mul : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
  val leq : ('c,int,int) repr -> ('c,int,int) repr -> ('c,bool,bool) repr
  val if_ : ('c,bool,bool) repr -> (unit -> 'x) -> (unit -> 'x) -> (('c,'sa,'da) repr as 'x)

end
\end{code}
\end{floatrule}
\caption{A (Meta)OCaml embedding of our object language that supports partial evaluation}
\label{fig:ocaml}
\end{figure*}

The interpreters |R|, |L| and |C| above only relied on the last
argument of the |repr| type, which is unchanged. And so is their
implementation; the newly added argument |'sv| will be phantom.
For example, we would now write the interpreter |C| as
\begin{code}
module C = struct
  type ('c,'sv,'dv) repr = ('c,'dv) code
  (* the rest is the same *)
\end{code}
When writing the partial evaluator however, we make use of the
type argument |'sv|. Here's the corresponding instance of the
Symantics:
